### 2019-02-25 implementation.tex
full TSTL example

\begin{figure}
\begin{code}
@import storage
\vspace{0.1in}
init: storage.init()
\vspace{0.1in}
pool: <int> 5
\vspace{0.1in}
<int> := <[-10..10]>
<int> := storage.getNewVal()
\{KeyError,ValueError\} storage.store(<int>)
\vspace{0.1in}
property: len(storage.contents()) == len(set(storage.contents()))
\end{code}
\caption{TSTL harness for Python library with (independent) vertical and
  horizonal nondeterminisms}
\label{fig:simpleexample}
\end{figure}

@@

Figures \ref{fig:simpleexample} and \ref{fig:simpleexamplecode} show a self-contained example of a
Python library and test harness with both vertical and horizontal nondeterminism.

>>>>

\begin{figure}
{\scriptsize
\begin{code}
@from storage import *
\vspace{0.1in}
init: init()
\vspace{0.1in}
pool: <int> 5
\vspace{0.1in}
<int> := <[-10..10]>
<int> := getNewVal()
\{KeyError,ValueError\} store(<int>)
\vspace{0.1in}
property: len(contents()) == len(set(contents()))
\end{code}
}
\caption{TSTL harness for Python library with (independent) vertical and
  horizonal nondeterminisms}
\label{fig:simpleexample}
\end{figure}

@@

Figures \ref{fig:simpleexample} and \ref{fig:simpleexamplecode} show a self-contained example of a
Python library and test harness with both vertical and horizontal
nondeterminism.  The example is not meant to be realistic, but to
better concretely explain the detection and debugging of
nondeterminism.  A few additional TSTL features need to be described.
First, lines beginning with an {\tt @} indicate raw Python code, here
used to import the module to be tested.  Second, {\tt init:} indicates
code that is to be executed at the beginning of each test, in
order to restore the system under test to its initial state.

\subsection{Horizontal Nondeterminism}

If we compile and test this code without using any of the
nondeterminism-detection functionality, TSTL will not report any
failing tests.  The property (avoidance of duplicate entries in
storage) holds:

{\scriptsize
\begin{code}
 \$ tstl storage.tstl
 ...
 \$ tstl\_rt
 Random testing using config=Config...
 storage: TEST \#1396 STEP \#0 (0:00:30.000188) Mon Feb
 25 11:13:18 2019 [ 15 stmts 22 branches ] (no cov+ for
 1391 tests) 139500 TOTAL ACTIONS (4649.97/s) (test
 0.0/s) 2 cov+ tests
\end{code}
}

TSTL reports information on the testing in a status line, here showing
that it has performed 1,396 tests, with no failures, and two tests
that increased code coverage.  TSTL can be run in a mode that stores
minimized versions of the tests that obtained new coverage, for use
as, e.g., regressions:

{\scriptsize
\begin{code}
 \$ tstl storage.tstl
 ...
 \$ tstl\_rt --quickTests
\end{code}
}

\begin{figure}
{\bf quick.0.test:}
\begin{code}
self.p\_int[0] = getNewVal() 
store(self.p\_int[0]) 
store(self.p\_int[0]) 
self.p\_int[0] = -10 
store(self.p\_int[0]) 
\end{code}
\vspace{0.1in}
{\bf quick.1.test:}
\begin{code}
self.p\_int[0] = -8
store(self.p\_int[0]) 
self.p\_int[0] = getNewVal() 
\end{code}
\caption{Coverage maximizing tests for Python library with (independent) vertical and
  horizonal nondeterminisms}
\label{fig:simpleexampletests}
\end{figure}

This will result in generation of two\footnote{In some rare cases,
  TSTL may generate one test that achieves complete code coverage, but
  this will only happen rarely here.} files, {\tt quick.0.test} and
{\tt quick.1.test}.  Sometimes, TSTL will fail at this point with an internal error
({\tt REDUCED TEST DOES NOT PRESERVE COVERAGE}), due to the attempt to reduce one of the coverage-obtaining
tests losing coverage, which is a consequence of horizontal
nondeterminism.  However, TSTL will often produce
test files such as those shown in Figure
\ref{fig:simpleexampletests}.  TSTL lets us convert these tests, which
are in TSTL's internal format, into standalone Python regression tests
that check the properties in the test harness and ensure that the
behavior (values of pools) are the same as in the recorded run:

{\scriptsize
\begin{code}
 \$ tstl\_standalone quick.0.test quick0.py --regression
 ...
 \$ tstl\_standalone quick.1.test quick1.py --regression
\end{code}
}

However, if we execute these Python tests, they will be very flaky;
sometimes they will pass, but {\tt quick0.py} will almost always fail, and
{\tt quick1.py} will fail about two thirds of the time.  The reason,
of course, is a trivial case of horizontal nondeterminism, due to the
use of the {\tt random.randint} function in {\tt getNewValue}.  The
regression test will record whatever value {\tt getNewValue} returned
the when the TSTL test file was executed in order to produce the
Python standalone test.  The value in future executions will seldom
match this value.  In this case, removing the nondeterminism is easy
(simply add a fixed random seed to the {\tt init:} line in the
harness, e.g., {\tt init: init(); random.seed(0)}).  However, in
real-world testing, the source of the nondeterminism, may not be
obvious, and only some regression tests may (infrequently) fail.
However, if we ask the random tester to check for nondeterminism, it
quickly shows us what the problem is:

{\scriptsize
\begin{code}
 \$ tstl\_rt --checkDeterminism
 ...
 int2 = getNewVal()              \# STEP 50
 MISMATCH IN REPLAY VALUE:
    int2 : 7 VS. -5
 TEST WAS NOT DETERMINISTIC!  WRITING FAILURE AS
 ndfail.test
\end{code}
}

We can use TSTL's test reduction tool to produce a small test showing
the problem:

{\scriptsize
\begin{code}
 \$ tstl\_reduce ndfail.test hnd.test --checkDeterminism
 ...
 NEW LENGTH 1
 int0 = getNewVal()              \# STEP 0

 TEST WRITTEN TO hnd.test
\end{code}
}

Nothing is required to produce nondeterminism other than calling {\tt
  getNewVal}, and this test is more frequently nondeterministic than
the original, since it calls {\tt getNewVal} without any stored data
to restrict the range of random values.

\subsection{Vertical Nondeterminism}

The {\tt storage.py} program also contains a (again, unrealistically
simple) example of \emph{vertical} nondeterminism.  The {\tt store}
method is supposed to 1) allow only a single copy of a value to be in
the storage and 2) to disallow negative values.  Unfortunately, the
check for negatives is \emph{after} the value is stored, which means
that a second attempt to store the same negative value will result in
a {\tt KeyError}, not a {\tt ValueError}.  TSTL can detect this
problem if we compile the harness with {\tt
  --checkFailureDeterminism}, which produces a version of the testing
interface that ensures failing actions (ones that raise an expected
exception) always fail again if performed again without any
intervening actions, and always fail with the same exception.  Note
that the check for the same exception is a slight extension of our formal
definition of failure determinism, recording the last observed
exception as an additional visible state element.

{\scriptsize
\begin{code}
 \$ tstl storage.tstl --checkFailureDeterminsim
 ...
 \$ tstl\_rt
 ...
 FINAL VERSION OF TEST, WITH LOGGED REPLAY:
 int0 = -4                                   \# STEP 0
 ACTION: int0 = -4 
 int0 = None : <type 'NoneType'>
 => int0 = -4 : <type 'int'>
 ==================================================
 store(int0)                                 \# STEP 1
 ACTION: store(int0) 
 int0 = -4 : <type 'int'>
 RAISED EXPECTED EXCEPTION:
   <type 'exceptions.ValueError'> 
 RAISED EXCEPTION:
   <type 'exceptions.KeyError'>  ON RETRY
   ERROR: (<type 'exceptions.AssertionError'>,
   AssertionError('Failure with different exception.',),
 <traceback object at 0x105e9edd0>)
\end{code}
}

With failure nondeterminism, we don't have to use TSTL's specialized reduction
tool, since the failure does not require multiple replays of a test.
The random tester itself can properly reduce the test case, since it
is just a ``normal'' failure, exhibited by a single trace.
